Supplementary Information

### 2021-08-27 doc/manuscript/ForestGEO_treerings.Rmd
added species authorities

SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* at SCBI."
Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* at Zofin."
CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* at Cedar Breaks."
SC_traditional_comparison_legend <- "Figure S4. Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* at Scotty Creek."

@@

![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-seasonal window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site. In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year. Example plots show raw data and partial effects of each variable from the best multivariate model for *Pinus ponderosa* at Little Tesuque, New Mexico, USA.](tables_figures/schematic_figure.png)

@@

To test for year effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
Specifically, we required that the species be represented by cores from  $\ge$ 3 trees and that the core record spanned $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm (exception: *Abies alba* at ZOF, where mature trees < 50 cm DBH are extremely rare).
Species that failed to meet these criteria (n= 8; Table `r sampling_details_table`) were excluded from the analysis of temporal trends, but were included in analyses of climate and DBH and their interactions.
We then ran models as described above, including a  first-order linear effect of year.
We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals that are not accounted for by DBH or year [@brienen_detecting_2012; @brienen_tree_2017]. 
To verify that GLS model trends for year were not an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade.

@@

```{r , eval=FALSE, echo=FALSE}
## Tallies for the paragraph below (Feb 11)
#BCNM: 1 species with both, 2 with just P
#HKK: 3 species with both, 1 with just T (CHTA)
#SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
#LDW:  4 species with both,  2  with just P (ACSA, CAOV)
#HF:  4 species with both
#ZOF:  3 species with both
#NIO:  1 species just T 
#LT:  2 species with both
#CB:  8 species with both
#SC:  1 species with both
```
### Climate responses {-#Climate}
Analyses of species-specific responses at each site used the GLS model to test for first- and negative second- order linear effects of both a precipitation and a temperature variable. 
Both a precipitation and a temperature variable were included in the top model for 78% (n=36 of 46) of site-species combinations (Fig. `r clim_sens`). 
There were seven site-species combinations for which only a precipitation term was significant (two at BCNM, three at SCBI, and two at LDW), two for which only a temperature term was significant (*Chukrasia tabularis* at HKK and *Betula papyrifera* at NIO), and none with no significant climatic effects on RW. 
Below, we summarize the precipitation and temperature variables included in these models and their direction of response.

@@

```{r , eval=TRUE, echo=FALSE}
###other:
n_ssc= 45 #number species-site combinations
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
percent_T_or_P_responses_nonlinear=(32+18)/(42+38)
```
Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species-site combinations, and were predominantly positive (Table `r hypotheses`, Fig. `r clim_sens`).
Specifically, there were positive first-order linear terms for precipitation for all but one species-site combination (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig. `r clim_sens`).

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r round(percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg)`%; `r round(percent_temperature_responses_1st_order_negative)`% with negative first-order term, `r round(percent_temperature_responses_1st_order_pos_2nd_order_neg)`% with positive first-order term but negative second-order term; Fig. `r clim_sens`).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with the most influential seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominantly positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).

@@

![**Figure 5 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution. Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)

@@

For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative). 
There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
This pattern was common among light-demanding species (6 of 13 site-species combinations; Tables `r hypotheses`, `r species_table`; e.g., *Melia azedarach* at HKK, *Populous tremuloides* at CB) and in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig. `r size`).
At the other end of the spectrum, some species had low RW at small DBH, increased to peak RW at intermediate DBH, and subsequently declined.
This pattern was common among shade-tolerant species (8 of 11 site-species combinations; Table `r hypotheses`; e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp. at SCBI and ZOF, *Picea* spp. at ZOF and CB; Table `r species_table`), and universal for shade-tolerant species in models accounting for calendar year (Fig. `r SummaryFigure`).

Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, typically increasing to a peak at intermediate DBH and then declining (Table `r hypotheses`, Fig. `r size`).
Best models for BAI included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
For $\Delta$AGB, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for DBH and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).

![**Figure 6 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Models presented here include climate variables and DBH as fixed effects. Relationships for species are plotted when included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)

@@

There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species-site combinations tested (Figs. `r year_fig`, `r year_trends_all`).
In `r percent_neg_of_sig_trends` of cases (depending on growth metric), the growth trend over time was negative.
Declines were particularly prevalent in secondary or disturbed forests, occurring in `r round(n_best_model_BAI_young/n_tested_young_forest*100)`% of species-site combinations (`r round(n_neg_BAI_young/n_best_model_BAI_young*100)`% of all species with significant year effects) at the seven disturbed sites (Table `r hypotheses`, Fig. `r year_fig`).
In older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Figs. `r year_fig`, `r year_trends_all`).
Significant positive growth trends were observed for only three species (consistently across all three growth metrics), *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* at HKK, *Pinus strobiformis* at LT, and *Picea engelmannii* and *Pinus longaeva* at CB.

@@

![**Figure 7 | Effect of year, when included in the best model, on BAI.**  Models presented here include climate variables, DBH, and year as fixed effects. For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. ](tables_figures/Year_responses_BAI_only.png)

@@

There was substantial variation across species-site combinations in the population mean relationship between DBH and growth rate (Fig. `r size`).
Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
This variation was driven by two primary, interrelated factors: species ecology and stand history.
As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`), particularly when the effects of calendar year were accounted for in the model (Fig. `r SummaryFigure`). 
However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
These results imply that while species that typically grow in high-light conditions commonly display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), average RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018]. 
Our finding that on average across populations, BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that mean $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models reduced the frequency of RW declines with DBH (Fig. `r SummaryFigure`) and tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the growth declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

@@

Within the three older forests (ZOF, LT, CB), population-mean growth trends were mixed (Table `r hypotheses`, Figs. `r SummaryFigure`, `r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s, consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008]. 
Nonlinear trends such as this would be more accurately described by a nonlinear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
Notably, there were only three species -- *Fagus sylvatica* at ZOF and *Picea pungens* and *Pinus flexilis* at CB -- whose population-mean growth rate increased significantly over the analysis time frame (Fig. `r year_fig`).

>>>>

SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* L. at SCBI."
Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* Mill. at Zofin."
CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* (Mirb.) Franco at Cedar Breaks."
SC_traditional_comparison_legend <- "Figure S4. Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* a (Mill.) Britton, Sterns and Poggenb. at Scotty Creek."

@@

![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-seasonal window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site. In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year. Example plots show raw data and partial effects of each variable from the best multivariate model for *Pinus ponderosa* P. Lawson & C. Lawson at Little Tesuque, New Mexico, USA.](tables_figures/schematic_figure.png)

@@

To test for year effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
Specifically, we required that the species be represented by cores from  $\ge$ 3 trees and that the core record spanned $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm (exception: *Abies alba* Mill. at ZOF, where mature trees < 50 cm DBH are extremely rare).
Species that failed to meet these criteria (n= 8; Table `r sampling_details_table`) were excluded from the analysis of temporal trends, but were included in analyses of climate and DBH and their interactions.
We then ran models as described above, including a  first-order linear effect of year.
We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals that are not accounted for by DBH or year [@brienen_detecting_2012; @brienen_tree_2017]. 
To verify that GLS model trends for year were not an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade.

@@

```{r , eval=FALSE, echo=FALSE}
## Tallies for the paragraph below (Feb 11)
#BCNM: 1 species with both, 2 with just P
#HKK: 3 species with both, 1 with just T (CHTA)
#SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
#LDW:  4 species with both,  2  with just P (ACSA, CAOV)
#HF:  4 species with both
#ZOF:  3 species with both
#NIO:  1 species just T 
#LT:  2 species with both
#CB:  8 species with both
#SC:  1 species with both
```
### Climate responses {-#Climate}
Analyses of species-specific responses at each site used the GLS model to test for first- and negative second- order linear effects of both a precipitation and a temperature variable. 
Both a precipitation and a temperature variable were included in the top model for 78% (n=36 of 46) of site-species combinations (Fig. `r clim_sens`). 
There were seven site-species combinations for which only a precipitation term was significant (two at BCNM, three at SCBI, and two at LDW), two for which only a temperature term was significant (*Chukrasia tabularis*  A. Juss. at HKK and *Betula papyrifera* Marshall at NIO), and none with no significant climatic effects on RW. 
Below, we summarize the precipitation and temperature variables included in these models and their direction of response.

@@

```{r , eval=TRUE, echo=FALSE}
###other:
n_ssc= 45 #number species-site combinations
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
percent_T_or_P_responses_nonlinear=(32+18)/(42+38)
```
Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species-site combinations, and were predominantly positive (Table `r hypotheses`, Fig. `r clim_sens`).
Specifically, there were positive first-order linear terms for precipitation for all but one species-site combination (*Tsuga canadensis*  (L.) Carrière at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* Britton at HF; Fig. `r clim_sens`).

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r round(percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg)`%; `r round(percent_temperature_responses_1st_order_negative)`% with negative first-order term, `r round(percent_temperature_responses_1st_order_pos_2nd_order_neg)`% with positive first-order term but negative second-order term; Fig. `r clim_sens`).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with the most influential seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* (Mill.) Britton, Sterns & Poggenb. responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominantly positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).

@@

![**Figure 5 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution. Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD. Species authorities are given in Table S2.](tables_figures/pre_temp_groups_dbh_interactions.png)

@@

For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative). 
There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
This pattern was common among light-demanding species (6 of 13 site-species combinations; Tables `r hypotheses`, `r species_table`; e.g., *Melia azedarach* L. at HKK, *Populous tremuloides* Michx. at CB) and in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig. `r size`).
At the other end of the spectrum, some species had low RW at small DBH, increased to peak RW at intermediate DBH, and subsequently declined.
This pattern was common among shade-tolerant species (8 of 11 site-species combinations; Table `r hypotheses`; e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp. at SCBI and ZOF, *Picea* spp. at ZOF and CB; Table `r species_table`), and universal for shade-tolerant species in models accounting for calendar year (Fig. `r SummaryFigure`).

Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, typically increasing to a peak at intermediate DBH and then declining (Table `r hypotheses`, Fig. `r size`).
Best models for BAI included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* L. at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* P. Lawson & C. Lawson at LT and *Pinus longaeva*  D.K. Bailey at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
For $\Delta$AGB, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for DBH and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).

![**Figure 6 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Models presented here include climate variables and DBH as fixed effects. Relationships for species are plotted when included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. Species authorities are given in Table S2.](tables_figures/DBH_responses.png)

@@

There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species-site combinations tested (Figs. `r year_fig`, `r year_trends_all`).
In `r percent_neg_of_sig_trends` of cases (depending on growth metric), the growth trend over time was negative.
Declines were particularly prevalent in secondary or disturbed forests, occurring in `r round(n_best_model_BAI_young/n_tested_young_forest*100)`% of species-site combinations (`r round(n_neg_BAI_young/n_best_model_BAI_young*100)`% of all species with significant year effects) at the seven disturbed sites (Table `r hypotheses`, Fig. `r year_fig`).
In older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Figs. `r year_fig`, `r year_trends_all`).
Significant positive growth trends were observed for only three species (consistently across all three growth metrics), *Fagus sylvatica* L. at ZOF, *Picea pungens* Engelm. and *Pinus flexilis*  E. James at CB, and all were modest compared to the steep negative trends observed for some species.
Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* A. Juss. at HKK, *Pinus strobiformis* Engelm. at LT, and *Picea engelmannii* Engelm. and *Pinus longaeva* at CB.

@@

![**Figure 7 | Effect of year, when included in the best model, on BAI.**  Models presented here include climate variables, DBH, and year as fixed effects. For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. Species authorities are given in Table S2. ](tables_figures/Year_responses_BAI_only.png)

@@

There was substantial variation across species-site combinations in the population mean relationship between DBH and growth rate (Fig. `r size`).
Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
This variation was driven by two primary, interrelated factors: species ecology and stand history.
As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`), particularly when the effects of calendar year were accounted for in the model (Fig. `r SummaryFigure`). 
However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa*  (Kurz) Craib at HKK, *Liriodendron tulipifera* L. at LDW).
These results imply that while species that typically grow in high-light conditions commonly display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), average RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018]. 
Our finding that on average across populations, BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that mean $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models reduced the frequency of RW declines with DBH (Fig. `r SummaryFigure`) and tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the growth declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

@@

Within the three older forests (ZOF, LT, CB), population-mean growth trends were mixed (Table `r hypotheses`, Figs. `r SummaryFigure`, `r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides*  Michx. at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s, consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008]. 
Nonlinear trends such as this would be more accurately described by a nonlinear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
Notably, there were only three species -- *Fagus sylvatica* at ZOF and *Picea pungens* and *Pinus flexilis* at CB -- whose population-mean growth rate increased significantly over the analysis time frame (Fig. `r year_fig`).
